---
title: "R Notebook"
output: html_notebook
---


```{r}
#install.packages("Metrics")
#install.packages("caret")
#install.packages("readx")
#install.packages("reshape2")
#install.packages("ggpubr")
#install.packages("stringr")
#install.packages("skimr")
#install.packages("e1071")
#install.packages("pROC")
library(Metrics)
library(caret)
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)
library(skimr)
library(tidyr)
library(reshape2)
library(ggpubr)
library(stringr)
library(e1071)
library(pROC)
```



```{r}
happy15_df = read.csv ("2015.csv")
happy16_df = read.csv ("2016.csv")
happy17_df = read.csv ("2017.csv")
happy18_df = read.csv ("2018.csv")
happy19_df = read.csv ("2019.csv")
head(happy19_df)
head(happy18_df)
head(happy17_df)
head(happy16_df)
head(happy15_df)
```


#Change year 2018 datasets column names base year 2017 datasets
```{r}
happy18_df=plyr::rename(happy18_df, replace = c( "Country.or.region"="Country", 
                                  "Overall.rank"="Happiness.Rank" ,
                                  "GDP.per.capita"="Economy..GDP.per.Capita.",
                                  "Healthy.life.expectancy"="Health..Life.Expectancy.",
                                  "Freedom.to.make.life.choices"="Freedom",
                                  "Perceptions.of.corruption"="Trust..Government.Corruption.",
                                  "Social.support"="Family",
                                  "Score"="Happiness.Score"))
colnames(happy18_df)
```


#Change year 2019 datasets column names base year 2017 datasets
```{r}
happy19_df=plyr::rename(happy19_df, replace = c( "Country.or.region"="Country", 
                                  "Overall.rank"="Happiness.Rank" ,
                                  "GDP.per.capita"="Economy..GDP.per.Capita.",
                                  "Healthy.life.expectancy"="Health..Life.Expectancy.",
                                  "Freedom.to.make.life.choices"="Freedom",
                                  "Perceptions.of.corruption"="Trust..Government.Corruption.",
                                  "Social.support"="Family",
                                  "Score"="Happiness.Score"))
colnames(happy19_df)
```


#Insert year column at first position (index 0)
```{r}
happy15_df<-cbind(Year=2015,happy15_df)

happy16_df<-cbind(Year=2016,happy16_df)

happy17_df<-cbind(Year=2017,happy17_df)

happy18_df<-cbind(Year=2018,happy18_df)

happy19_df<-cbind(Year=2019,happy19_df)
head(happy19_df)
head(happy18_df)
head(happy17_df)
head(happy16_df)
head(happy15_df)
```




#Change column type for emerging dataset
```{r}
happy18_df$Trust..Government.Corruption. = as.numeric(happy18_df$Trust..Government.Corruption.)

str(happy18_df)

```



#Merge data from 2015-2019
```{r}
happy15_16<-dplyr::bind_rows(happy15_df,happy16_df)

happy15_16_17<-dplyr::bind_rows(happy15_16,happy17_df)

happy18_19<-dplyr::bind_rows(happy18_df,happy19_df)

df<-dplyr::bind_rows(happy18_19,happy15_16_17)

head(df)
```




#Change Happiness.Rank data type
```{r}
df$Happiness.Rank  = as.numeric(df$Happiness.Rank )

str(df)
```





#Remove unnescessary columns
```{r}
#Count NA value in every column
colSums(is.na(df))
```


#Now remove the unnecessary columns
```{r}
df = subset(df, select = -c(Lower.Confidence.Interval,Upper.Confidence.Interval,Dystopia.Residual,Standard.Error,Whisker.high,Whisker.low))

colSums(is.na(df))
```



```{r}
df$Trust..Government.Corruption.[is.na(df$Trust..Government.Corruption.)] <- median(df$Trust..Government.Corruption., na.rm = T)
colSums(is.na(df))
```


```{r}
aggregate(df$Country, by=list(df$Year), FUN=length)
```


```{r}
Country_2015 = subset(df, Year == 2015)$Country
Country_2016 = subset(df, Year == 2016)$Country
Country_2017 = subset(df, Year == 2017)$Country
Country_2018 = subset(df, Year == 2018)$Country
Country_2019 = subset(df, Year == 2019)$Country
```




```{r}
common_country =intersect(intersect(intersect(intersect(Country_2015,
Country_2016),Country_2017),Country_2018),Country_2019)
length(common_country)
```

```{r}
df1 = subset(df,Country %in% common_country)
print(paste("The amount of rows in the dataset is: ",dim(df1)[1]))
print(paste("The amount of columns in the dataset is: ",dim(df1)[2]))
```



```{r}
common_region <- unique(subset(df1, Region!="NA", c(Country, Region)))

head(common_country)
common_region
```




```{r}
assign_region <- function(x){
  Region <- common_region$Region[common_region$Country == x]
}

for(country in common_country)
      df1$Region[df1$Country == country] <- assign_region(country)
```



```{r}
colSums(is.na(df1))
```



```{r}
skimr::skim_without_charts(df1)
```

```{r}
str(df1)
```


```{r}
df1 %>%
  filter(Year == 2015) %>%
  arrange(-Happiness.Score) %>%
  slice_head(n=10) %>%
  ggplot(aes(reorder(Country, Happiness.Score), Happiness.Score)) +
  geom_point(colour = "red", size = 3) +
  theme(text=element_text(size=10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2015", x = "")
```


```{r}
# Filter data for the year 2015
df_top_10 <- df1 %>%
  filter(Year == 2015) %>%
  arrange(desc(Happiness.Score)) %>%
  slice_head(n = 10)

# Plotting
ggplot(df_top_10, aes(reorder(Country, Happiness.Score), Happiness.Score, size = Happiness.Score)) +
  geom_point(colour = "red") +
  theme(text = element_text(size = 10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2015", x = "") +
  scale_size_continuous(range = c(3, 10))
```



```{r}
df1 %>%
  filter(Year == 2016) %>%
  arrange(-Happiness.Score) %>%
  slice_head(n=10) %>%
  ggplot(aes(reorder(Country, Happiness.Score), Happiness.Score)) +
  geom_point(colour = "red", size = 3) +
  theme(text=element_text(size=10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2016", x = "")
```




```{r}
# Filter data for the year 2015
df_top_10 <- df1 %>%
  filter(Year == 2016) %>%
  arrange(desc(Happiness.Score)) %>%
  slice_head(n = 10)

# Plotting
ggplot(df_top_10, aes(reorder(Country, Happiness.Score), Happiness.Score, size = Happiness.Score)) +
  geom_point(colour = "red") +
  theme(text = element_text(size = 10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2016", x = "") +
  scale_size_continuous(range = c(3, 10))
```



```{r}
df1 %>%
  filter(Year == 2017) %>%
  arrange(-Happiness.Score) %>%
  slice_head(n=10) %>%
  ggplot(aes(reorder(Country, Happiness.Score), Happiness.Score)) +
  geom_point(colour = "red", size = 3) +
  theme(text=element_text(size=10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2017", x = "")
```


```{r}
# Filter data for the year 2015
df_top_10 <- df1 %>%
  filter(Year == 2017) %>%
  arrange(desc(Happiness.Score)) %>%
  slice_head(n = 10)

# Plotting
ggplot(df_top_10, aes(reorder(Country, Happiness.Score), Happiness.Score, size = Happiness.Score)) +
  geom_point(colour = "red") +
  theme(text = element_text(size = 10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2017", x = "") +
  scale_size_continuous(range = c(3, 10))
```



```{r}
df1 %>%
  filter(Year == 2018) %>%
  arrange(-Happiness.Score) %>%
  slice_head(n=10) %>%
  ggplot(aes(reorder(Country, Happiness.Score), Happiness.Score)) +
  geom_point(colour = "red", size = 3) +
  theme(text=element_text(size=10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2018", x = "")
```


```{r}
# Filter data for the year 2015
df_top_10 <- df1 %>%
  filter(Year == 2018) %>%
  arrange(desc(Happiness.Score)) %>%
  slice_head(n = 10)

# Plotting
ggplot(df_top_10, aes(reorder(Country, Happiness.Score), Happiness.Score, size = Happiness.Score)) +
  geom_point(colour = "red") +
  theme(text = element_text(size = 10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2018", x = "") +
  scale_size_continuous(range = c(3, 10))
```




```{r}
df1 %>%
  filter(Year == 2019) %>%
  arrange(-Happiness.Score) %>%
  slice_head(n=10) %>%
  ggplot(aes(reorder(Country, Happiness.Score), Happiness.Score)) +
  geom_point(colour = "red", size = 3) +
  theme(text=element_text(size=10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2019", x = "")
```



```{r}
# Filter data for the year 2015
df_top_10 <- df1 %>%
  filter(Year == 2019) %>%
  arrange(desc(Happiness.Score)) %>%
  slice_head(n = 10)

# Plotting
ggplot(df_top_10, aes(reorder(Country, Happiness.Score), Happiness.Score, size = Happiness.Score)) +
  geom_point(colour = "red") +
  theme(text = element_text(size = 10)) + 
  coord_flip() +
  labs(title = "The 10 happiest countries in 2019", x = "") +
  scale_size_continuous(range = c(3, 10))
```



```{r}
# Group by Country and Year, calculate the mean happiness variables
Top10_Happiest <- df1 %>%
  select(Country, Year, Happiness.Score) %>%
  group_by(Country, Year) %>%
  summarise(Average_Happiness = mean(Happiness.Score, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(Average_Happiness)) %>%
  group_by(Year) %>%
  slice_head(n = 10)

# Create a bar plot showing the top 10 happiest countries for each year
ggplot(Top10_Happiest, aes(x = reorder(Country, Average_Happiness), y = Average_Happiness, fill = Year)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Happiest Countries for Different Years",
       x = "Country",
       y = "Average Happiness Score",
       fill = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



```{r}
Happiness.Continent <- df1 %>%
                          select(-c(Year,Happiness.Score,Happiness.Rank))%>%
                          group_by(Region) %>%
                          summarise_at(vars(-Country), funs(mean(., na.rm=TRUE)))


Happiness.Continent.melt <- melt(Happiness.Continent)


# Faceting
ggplot(Happiness.Continent.melt, aes(y=value, x=Region, color=Region, fill=Region)) + 
  geom_bar( stat="identity") +    
  facet_wrap(~variable) + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Average value of happiness variables for different regions", 
       y = "Average value") 

```





```{r}
df1 %>%
  group_by(Country) %>%
  summarise(mscore = mean(Happiness.Score)) %>%
  arrange(-mscore) %>%
  slice_head(n=10) %>%
  
  ggplot(aes(reorder(Country, mscore), mscore)) +
  geom_point() +
  theme_bw() +
  
  coord_flip() +
  labs(title = "Happiness Score by Country",
       x = "", y = "Average happiness score")
```



```{r}
df1 %>%
  group_by(Country) %>%
  summarise(mscore = mean(Happiness.Score, na.rm = TRUE)) %>%
  arrange(desc(mscore)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(reorder(Country, mscore), mscore)) +
  geom_point(colour = "red") +
  theme_bw() +
  coord_flip() +
  labs(title = "Happiness Score by Country",
       x = "", y = "Average happiness score")
```



```{r}
df1 %>%
  mutate(y = as.character(Year)) %>%
  select(y, Country, Region, Happiness.Score) %>%
  pivot_wider(names_from = y, values_from = Happiness.Score, names_prefix = "y_") %>%
  mutate(p = (y_2019 - y_2015) / y_2015 * 100) %>%
  arrange(desc(p)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(reorder(Country, p), p)) +
  geom_point(colour = "red") +
  theme_bw() +
  coord_flip() +
  labs(title = "The 10 most progressive countries from 2015 - 2019",
       y = "Percentage Increase of Happiness Score", x = "")
```



```{r}
df1 %>%
  group_by(Country) %>%
  summarise(gdp = mean(Economy..GDP.per.Capita.),
            family = mean(Family),
            life.expectancy = mean(Health..Life.Expectancy.),
            freedom = mean(Freedom),
            generosity = mean(Generosity),
            corruption = mean(Trust..Government.Corruption.)) %>%
  pivot_longer(c(gdp, family, life.expectancy, freedom, generosity, corruption),
               names_to = "factor", values_to = "value") %>%
  ggplot(aes(reorder(factor, value), value)) +
  geom_bar(stat = "identity", fill = "darkgreen", width = 0.55, alpha = 0.7) +
  geom_text(aes(label = paste0(round(value, 2)), vjust = -0.5)) +
  theme_bw() +
  labs(title = "The mean value of factors by country", y = "", x = "")

```



```{r}
df1 %>%
  group_by(Region) %>%
  summarise(gdp = mean(Economy..GDP.per.Capita.),
            family = mean(Family),
            life.expectancy = mean(Health..Life.Expectancy.),
            freedom = mean(Freedom),
            generosity = mean(Generosity),
            corruption = mean(Trust..Government.Corruption.)) %>%
  pivot_longer(c(gdp, family, life.expectancy, freedom, generosity, corruption),
               names_to = "factor", values_to = "value") %>%
  ggplot(aes(reorder(factor, value), value)) +
  geom_bar(stat = "identity", fill = "darkgreen", width = 0.55, alpha = 0.7) +
  geom_text(aes(label = paste0(round(value, 2)), vjust = -0.5)) +
  theme_bw() +
  labs(title = "The mean value of factors by region", y = "", x = "")

```


```{r}
df1 %>%
  summarise(gdp = mean(Economy..GDP.per.Capita.),
            family = mean(Family),
            life.expectancy = mean(Health..Life.Expectancy.),
            freedom = mean(Freedom),
            generosity = mean(Generosity),
            corruption = mean(Trust..Government.Corruption.)) %>%
  pivot_longer(c(gdp, family, life.expectancy,freedom,generosity, corruption),
               names_to = "f", values_to = "value") %>%
  ggplot(aes(reorder(f, value), value)) +
  geom_bar(stat = "identity", fill = "red", width = 0.55, alpha = 0.7) +
  geom_text(aes(label = paste0(round(value, 2)), vjust = -0.5)) +
  theme_bw() +
  labs(title = "The mean value of the factors" , y = "", x = "")
```





#Find correlation using heatmap
# Drop Year,Country,Happiness.Rank,Region column before computing the heatmap.
```{r}
dataset = select(df1,-c("Year","Country","Happiness.Rank","Region"))
head(dataset)
```


```{r}
#install.packages("GGally")
library(GGally)

ggcorr(dataset, label = TRUE, label_round = 2, label_size = 3.5, size = 2, hjust = .85) +
  ggtitle("Correlation Heatmap") +
  theme(plot.title = element_text(hjust = 0.5))
```




#Regression models
# Split into train set (80%) and test set (20%)
```{r}
set.seed(123) 
split=0.80
trainIndex <- createDataPartition(dataset$Happiness.Score, p=split, list=FALSE) 
data_train <- dataset[ trainIndex,] 
data_test <- dataset[-trainIndex,]
```




#Train Multiple Linear Regression model with data_train
```{r}
# Fitting Multiple Linear Regression to the Training set
lm_model = lm(formula = Happiness.Score ~ .,
               data = data_train)

summary(lm_model)
```


# Predict happiness score with data_test
```{r}
y_pred_lm = predict(lm_model, newdata = data_test)
Actual_lm = data_test$Happiness.Score

Pred_Actual_lm <- as.data.frame(cbind(Prediction = y_pred_lm, Actual = Actual_lm))


gg.lm <- ggplot(Pred_Actual_lm, aes(Actual, Prediction )) +
  geom_point() + theme_bw() + geom_abline() +
  labs(title = "Multiple Linear Regression", x = "Actual happiness score",
       y = "Predicted happiness score") +
  theme(plot.title = element_text(family = "Helvetica", face = "bold", size = (15)), 
        axis.title = element_text(family = "Helvetica", size = (10)))
gg.lm
```
```{r}
data.frame(
  R2 = R2(y_pred_lm, data_test$Happiness.Score),
  RMSE = RMSE(y_pred_lm, data_test$Happiness.Score),
  MAE = MAE(y_pred_lm, data_test$Happiness.Score)
)
```






#Train Random Forest Regression model with data_train
```{r}

#install.packages("randomForest")
library(randomForest)

x_train_rf<-select(dataset,-c("Happiness.Score"))

          
set.seed(1234)
regressor_rf = randomForest(x = x_train_rf,
                         y = dataset$Happiness.Score,
                         ntree = 500)
```




#Predict happiness score with data_test
```{r}
# Predicting happiness score with Random Forest Regression
y_pred_rf = predict(regressor_rf, newdata = data_test)

Pred_Actual_rf <- as.data.frame(cbind(Prediction = y_pred_rf, Actual = data_test$Happiness.Score))


gg.rf <- ggplot(Pred_Actual_rf, aes(Actual, Prediction )) +
  geom_point() + theme_bw() + geom_abline() +
  labs(title = "Random Forest Regression", x = "Actual happiness score",
       y = "Predicted happiness score") +
  theme(plot.title = element_text(family = "Helvetica", face = "bold", size = (15)), 
        axis.title = element_text(family = "Helvetica", size = (10)))
gg.rf
```


```{r}
# Predicting happiness score with Random Forest Regression
y_pred_rf = predict(regressor_rf, newdata = data_test)

# Creating a data frame with predictions and actual values
Pred_Actual_rf <- as.data.frame(cbind(Prediction = y_pred_rf, Actual = data_test$Happiness.Score))

# Plotting the predictions vs actual values
gg.rf <- ggplot(Pred_Actual_rf, aes(Actual, Prediction)) +
  geom_point(colour = "red") +  # Set points color to red
  geom_abline(colour = "red") + # Set line color to red
  theme_bw() +
  labs(title = "Random Forest Regression", x = "Actual happiness score", y = "Predicted happiness score") +
  theme(plot.title = element_text(family = "Helvetica", face = "bold", size = 15), 
        axis.title = element_text(family = "Helvetica", size = 10))

# Display the plot
gg.rf
```



```{r}
data.frame(
  R2 = R2(y_pred_rf, data_test$Happiness.Score),
  RMSE = RMSE(y_pred_rf, data_test$Happiness.Score),
  MAE = MAE(y_pred_rf, data_test$Happiness.Score)
)
```






#Model Evaluation
```{r}
ggarrange(gg.lm,gg.rf, ncol = 2, nrow = 2)
```



```{r}
tc <- trainControl(method = "repeatedcv", 
                   number=10,#10-fold cross validation 
                   classProbs = TRUE,
                   savePredictions = TRUE, 
                   repeats = 3,
                   ## Estimate class probabilities
                   summaryFunction = multiClassSummary,)
```


```{r}
set.seed(123)
model_knn <- train(
  Happy.Level~., 
  data=data_train, 
  trControl=tc,
  preProcess = c("center","scale"),
  method="knn",
  metric='Accuracy',
  tuneLength=20
  ) 

model_knn
```



```{r}
pred_knn <- predict(model_knn, data_test)

cm_knn<-confusionMatrix(pred_knn, data_test$Happy.Level)

cm_knn
```



```{r}
knn_importance <- varImp(model_knn)

# Create box plot of importance of variables
ggplot(data = knn_importance, mapping = aes(x = knn_importance[, 1])) + 
  geom_boxplot(fill = "red", color = "red") + # Create box plot with red color
  labs(title = "Variable importance: K-Nearest Neighbours") + # Title
  theme_light() # Theme
```

